---
title: "Project 3 Modeling"
author: "Susan Hajmohammad"
format: html
---

## Introduction: 

After doing an exploratory data analysis on four variables in relation to diabetic outcomes, I found I was most interested to see how Sex, Income and Education performed as predictors of Diabetes diagnosis.  These were the four variables that, at a glance, showed possibly some effect of amount of diabetes. I also find it interesting that these predictors are not comorbidities, they are rather biological (sex) or socio-economic (income and education) categories that people fall into.  This means that without having extra medical measurements done, providers could be aware of how the risk differs between patients, and perhaps patients would consent to early screenings, etc.  Let's see how good these variables are at predicting a diabetic diagnosis!  

### Libraries: 

```{r}
#| warning: FALSE
#| message: FALSE

library(tidyverse)
library(ggplot2)
library(knitr)
library(tidymodels)
library(tree)
library(dplyr)
library(ranger)
```
### Data

```{r}
#read in diabetes data from folder
diabetes_data <- read_csv("diabetes_data.csv")

#now for chosen variables change each to factor and give meaningful labels for the levels
diabetes_data_reduced <- read_csv("diabetes_data.csv")%>% 
  mutate(Diabetes_binary = factor(Diabetes_binary, levels= c(0,1), labels=c("No Diabetes", "Diabetes")),
         Sex = factor(Sex, levels= c(0,1), labels = c("Female", "Male")), 
         # info sheet says 5 year increments
         Age = factor(Age, levels = c(1,2,3,4,5,6,7,8,9,10,11,12,13), labels = c("18to24", "25to29", "30to34", "35to39", "40to44", "45to49", "50to54", "55to59", "60to64", "65to69", "70to74", "75to79", "80plus")),
         Education= factor(Education, levels= c(1,2,3,4,5,6), labels=c("Never attended or only kindergarten","Elementary","Some high school", "High school graduate","Some college or technical school","College graduate")), 
         Income = factor(Income, levels = c(1,2,3,4,5,6,7,8), labels = c("< $10,000","< $15,000","< $20,000","< $25,000", "< $35,000","< $50,000","< $75,000","$75,000+")))%>%
  #select only relevant variables to work with smaller data set
  select(Diabetes_binary, Age, Sex, Education, Income)
```


### Data Split: 

```{r}
#split data into testing and training set with prop 70/30:
set.seed(415)
diab_split <- initial_split(diabetes_data_reduced, prop = 0.7)

test <- testing(diab_split)
  
train <- training(diab_split)

```

## Logistic Regression Models: 

A logistic regression model is ___.

### Recipes for LR models: 

```{r}
LR1_rec <- recipe(Diabetes_binary ~ Sex+ Income+ Education, data = train)%>%
  step_dummy(all_nominal_predictors())

LR2_rec  <- recipe(Diabetes_binary ~ Sex + Income, data = train) %>%
  step_dummy(all_nominal_predictors())

LR3_rec <- recipe(Diabetes_binary ~ Income + Education, data = train) %>%
  step_dummy(all_nominal_predictors())

```

### Model specs: 

```{r}
#logistic regression model specs: 

LR_spec <- logistic_reg() %>%
  set_engine("glm")

```

### Workflows: 

```{r}
LR1_wkf <- workflow() %>%
 add_recipe(LR1_rec) %>%
 add_model(LR_spec)
LR2_wkf <- workflow() %>%
 add_recipe(LR2_rec) %>%
 add_model(LR_spec)
LR3_wkf <- workflow() %>%
 add_recipe(LR3_rec) %>%
 add_model(LR_spec)
```


### CV 5-fold 

```{r}
diab_cv_folds <- vfold_cv(train, v= 5)
```

### Fit to our CV folds: 

```{r}
LR1_fit <- LR1_wkf %>%
 fit_resamples(diab_cv_folds, metrics = metric_set(accuracy, mn_log_loss))
LR2_fit <- LR2_wkf %>%
 fit_resamples(diab_cv_folds, metrics = metric_set(accuracy, mn_log_loss))
LR3_fit <- LR3_wkf %>%
 fit_resamples(diab_cv_folds, metrics = metric_set(accuracy, mn_log_loss))

```

### Collect metrics: 

```{r}
rbind(LR1_fit %>% collect_metrics(),
 LR2_fit %>% collect_metrics(),
 LR3_fit %>% collect_metrics()) %>%
 mutate(Model = c("Model1", "Model1", "Model2", "Model2", "Model3", "Model3")) %>%
 select(Model, everything())
```

The best model is model 1 because it is showing the lowest log loss and also has a smaller standard error.  It appears all of the models have similar accuracy so that didn't help in the selection process. 

### Best model test on test set: 

```{r}
final_LRmodel<- LR1_wkf |>
 last_fit(diab_split, metrics = metric_set(accuracy, mn_log_loss)) |>
 collect_metrics()

final_LRmodel
```

We got our best model, model 1, from the LR models and tested it on the test set.  On the test set our final LR model scored about an 86% accuracy and a log loss of 0.387. 

In summary, the mean CV log-loss was 0.387, 0.389 and 0.389 for models 1,2, and 3 respectively.  This means the best model at predicting diabetes outcomes with new data was model 1 which included the Sex, Education and Income predictors.  We will keep this model and keep these variables in as predictors! 


## Classification Tree: 

```{r}
#tree recipe 
tree_rec <- recipe(Diabetes_binary ~ Sex + Income + Education, data = train) %>%
  step_dummy(all_nominal_predictors())

#model spec
tree_mod <- decision_tree(
  cost_complexity = tune(), 
  min_n = 20) %>%
  set_engine("rpart") %>%
  set_mode("classification")

#tree workflow
tree_wkf <- workflow() %>%
  add_recipe(tree_rec) %>% 
  add_model(tree_mod)

# tree grid
cp_grid <- grid_regular(
  cost_complexity(),
  levels = 10)

# tune grid with log‑loss as the metric
temp <- tune_grid(
  tree_wkf,
  resamples = diab_cv_folds,
  grid      = cp_grid,
  metrics   = metric_set(mn_log_loss))

# get best tree, slice top value

best_tree <- temp %>%
  collect_metrics() %>%
  filter(.metric == "mn_log_loss") %>%
  arrange(mean) %>%
  slice(1)
#show best tree
best_tree

#finalize workflow
final_tree_wkf <- finalize_workflow(tree_wkf, best_tree)

#fit on full training set and evaluate on test set
tree_last <- last_fit(
  final_tree_wkf,
  diab_split,
  metrics = metric_set(accuracy, mn_log_loss))

tree_last %>% collect_metrics()

```
With these results, it looks our accuracy remains similar to our best LR model; however, the log-loss is higher (0.4033) than the best LR model.  We would not choose this model over the LR model 1 as it would give more confident predictions. 


## Random Forest

```{r}
#random forest spec
rf_spec <- rand_forest(mtry = tune(), trees = 10) %>%
 set_engine("ranger") %>%
 set_mode("classification")

#RF workflow 

rf_wkf <- workflow() %>%
 add_recipe(LR3_rec) %>%
 add_model(rf_spec)
 

#mtry grid from 1 to 3 because we have 3 predictors
rf_grid <- grid_regular(
  mtry(range = c(1, 3)), 
  levels = 7)


# fit to CV folds

rf_fit <- rf_wkf %>%
 tune_grid(resamples = diab_cv_folds,
 grid = rf_grid,
 metrics = metric_set(accuracy, mn_log_loss))

#look at log loss and sort

rf_fit %>%
 collect_metrics() %>%
 filter(.metric == "mn_log_loss") %>%
 arrange(mean)

#get best tuning parameter
rf_best_params <- select_best(rf_fit, metric = "mn_log_loss")
rf_best_params

#refit on whole traning set and test on testing set 
rf_final_wkf <- rf_wkf %>%
 finalize_workflow(rf_best_params)

rf_final_fit <- rf_final_wkf %>%
  last_fit(diab_split, metrics = metric_set(accuracy, mn_log_loss))

```


## Final Model Selection 

```{r}

final_LRmodel

tree_last %>% collect_metrics()

rf_final_fit %>% collect_metrics()

```


```{r}
#refit model without metrics for API
final_fit_result <- LR1_wkf |>
  last_fit(diab_split, metrics = metric_set(accuracy, mn_log_loss))

#extract workflow and save rds
fitted_workflow <- extract_workflow(final_fit_result)
saveRDS(fitted_workflow, "logRegFit_1.rds")

```

